Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-07-03 article-models-ttinternal.tex
reorganized to match content of article-models-ttinternal.tex

\begin{defn}
An \emph{extension homomorphism $\cfthom{f}$ from $\extalg{A}$ to
$\extalg{B}$ in context $\Gamma$} is a triple $\unfold{\cfthom{f}}$ consisting of
\begin{align*}
\jhom*
  {\Gamma}
  {\cftalgc{\cftalg{A}}}
  {\cftalgc{\cftalg{B}}}
  {\cfthomc{\cfthom{f}}}
  \\
\jfhom*
  {\Gamma}
  {\cftalgc{\cftalg{A}}}
  {\cftalgc{\cftalg{B}}}
  {\cfthomc{\cfthom{f}}}
  {\cftalgf{\cftalg{A}}}
  {\cftalgf{\cftalg{B}}}
  {\cfthomf{\cfthom{f}}}
  \\
\jfhom*
  {\Gamma}
  {{\cftalgc{\cftalg{A}}}{\cftalgf{\cftalg{A}}}}
  {{\cftalgc{\cftalg{B}}}{\cftalgf{\cftalg{B}}}}
  {\jvcomp{}{\cfthomc{\cfthom{f}}}{\cfthomf{\cfthom{f}}}}
  {\cftalgt{\cftalg{A}}}
  {\cftalgt{\cftalg{B}}}
  {\cfthomt{\cfthom{f}}}
\end{align*}
for which the diagrams
\begin{equation}\label{eq:exthom1}
\begin{tikzcd}
\ctxext{\cftalgc{\cftalg{A}}}{\cftalgf{\cftalg{A}}}
  \ar{r}{\jvcomp{}{\cfthomc{\cfthom{f}}}{\cfthomf{\cfthom{f}}}}
  \ar{d}[swap]{\cftctxext[\cftalg{A}]}
& \ctxext{\cftalgc{\cftalg{B}}}{\cftalgf{\cftalg{B}}}
  \ar{d}{\cftctxext[\cftalg{B}]}
  \\
\cftalgc{\cftalg{A}}
  \ar{r}[swap]{\cfthomc{\cfthom{f}}}
& \cftalgc{\cftalg{A}}
\end{tikzcd}
\end{equation}
and
\begin{equation}\label{eq:exthom2}
\begin{tikzcd}[column sep=huge]
\ctxext{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext[\cftalg{A}]}{\cftalgf{\cftalg{A}}}}
  \ar{r}{\jvcomp{}{\cfthomf{\cfthom{f}}}{\jcomp{}{\cftctxext[\cftalg{A}]}{\cfthomf{\cfthom{f}}}}}
  \ar{d}[swap]{\cftfamext[\cftalg{A}]}
& \jcomp{}{\cfthomc{\cfthom{f}}}{\ctxext{\cftalgf{\cftalg{B}}}{\jcomp{}{\cftctxext[\cftalg{B}]}{\cftalgf{\cftalg{B}}}}}
  \ar{d}{\jcomp{}{\cfthomc{\cfthom{f}}}{\cftfamext[\cftalg{B}]}}
  \\
\cftalgf{\cftalg{A}}
  \ar{r}[swap]{\cfthomf{\cfthom{f}}}
& \jcomp{}{\cfthomc{\cfthom{f}}}{\cftalgf{\cftalg{B}}}
\end{tikzcd}
\end{equation}
commute judgmentally.
\end{defn}

@@

@@

Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$. Weakening will be a
CFT-homomorphism
\begin{equation}\label{cftwk}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
  {\cftalg{F_{F_A}}}
  {\cftwk{\cftalg{A}}}
\end{equation}
satisfying judgmental equalities which express abstracted analogues of
the rules in sections \autoref{comp-ww,comp-ew,comp-0w}. The
definition of weakening algebras can be found in \autoref{sec:cftwkalg-defn}.

We begin with some trivial remarks to familiarize ourselves with the situation.
The morphism $\cftwk{\cftalg{A}}$ of \autoref{cftwk} is a triple
$(\cftwkc{\cftalg{A}},\cftwkf{\cftalg{A}},\cftwkt{\cftalg{A}})$ consisting of
the morphisms displayed in the following diagram:
\begin{equation*}
\begin{tikzcd}
\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftalgt{\cftalg{A}}}}
  \ar[fib]{d}
  \ar{r}{\cftwkt{\cftalg{A}}}
& \jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftalgt{\cftalg{A}}}}
  \ar[fib]{d}
  \\
\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftalgf{\cftalg{A}}}}
  \ar[fib]{d}
  \ar{r}{\cftwkf{\cftalg{A}}}
& \jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftalgf{\cftalg{A}}}}
  \ar[fib]{d}
  \\
\ctxwk{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{A}}}
  \ar{r}[swap]{\cftwkc{\cftalg{A}}}
& \jcomp{}{\cftctxext}{\cftalgf{\cftalg{A}}}
\end{tikzcd}
\end{equation*}

@@

\subsubsection{Weakening preserves itself}
Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$ and let
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
  {\jcomp{}{\cftctxext}{\cftalg{F_A}}}
  {\cfthom{w}}
\end{equation*}
be a CFT-homomorphism. The abstraction of the rules in \autoref{comp-ww} is the
condition that the diagram
\begin{equation}\label{diag:cftwk-cftwk}
\begin{tikzcd}[column sep=10em]
\ctxwk{\cftalgf{\cftalg{A}}}{{\cftalgf{\cftalg{F_A}}}{\cftalg{F_{F_A}}}}
  \ar{d}
    [swap]{ \ctxwk
        {{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}
        {\boldsymbol{\mathcal{F}}_{\cfthom{w}}}
      }
  \ar{r}{\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cfthom{w}}}}
& \ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_{F_{F_A}}}}
  \ar{d}{\jcomp{}{\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamext}}{\boldsymbol{\mathcal{F}}_{\cfthom{w}}}}
  \\
\ctxwk{{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}{\jcomp{}{\cfthomc{\cfthom{w}}}{\cftalg{F_{F_{F_A}}}}}
  \ar{r}[swap]{%
    \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\cfthom{w}}
                }
            }
        }
    }
& \jcomp{}{\cfthomc{\cfthom{w}}}{\cftalg{F_{F_{F_{F_A}}}}}
\end{tikzcd}
\end{equation}
of CFT-algebras in context 
$ \ctxext
    { { { {\Gamma}
          {\cftalgc{\cftalg{A}}}
          }
        { \cftalgf{\cftalg{A}}}
        }
      { \ctxwk
          {\cftalgf{\cftalg{A}}}
          {\cftalgf{\cftalg{A}}}
        }
      }
    { \ctxwk
        {\cftalgf{\cftalg{A}}}
        {\cftalgf{\cftalg{F_A}}}
      }$ %
commutes. In this diagram, it is not immediately obvious that the domain and
codomain of the CFT-homomorphism
$ \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\cfthom{w}}
                }
            }
        }$ %
from
$ \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
                }
            }
        }
$ to
$ \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\cftalg{F_{F_A}}}
                }
            }
        }
$ are indeed judgmentally equal to the indicated domain and codomain.

@@

>>>>

\begin{defn}
An \emph{extension homomorphism $\cfthom{f}$ from $\extalg{A}$ to
$\extalg{B}$ in context $\Gamma$} is a triple $\unfold{\cfthom{f}}$ consisting of
\begin{align*}
\jhom*
  {\Gamma}
  {\cftalgc{\cftalg{A}}}
  {\cftalgc{\cftalg{B}}}
  {\cfthomc{\cfthom{f}}}
  \\
\jfhom*
  {\Gamma}
  {\cftalgc{\cftalg{A}}}
  {\cftalgc{\cftalg{B}}}
  {\cfthomc{\cfthom{f}}}
  {\cftalgf{\cftalg{A}}}
  {\cftalgf{\cftalg{B}}}
  {\cfthomf{\cfthom{f}}}
  \\
\jfhom*
  {\Gamma}
  {{\cftalgc{\cftalg{A}}}{\cftalgf{\cftalg{A}}}}
  {{\cftalgc{\cftalg{B}}}{\cftalgf{\cftalg{B}}}}
  {\jvcomp{}{\cfthomc{\cfthom{f}}}{\cfthomf{\cfthom{f}}}}
  {\cftalgt{\cftalg{A}}}
  {\cftalgt{\cftalg{B}}}
  {\cfthomt{\cfthom{f}}}
\end{align*}
for which the diagrams
\begin{equation}\label{eq:exthom1}
\begin{tikzcd}
\ctxext{\cftalgc{\cftalg{A}}}{\cftalgf{\cftalg{A}}}
  \ar{r}{\jvcomp{}{\cfthomc{\cfthom{f}}}{\cfthomf{\cfthom{f}}}}
  \ar{d}[swap]{\cftctxext[\cftalg{A}]}
& \ctxext{\cftalgc{\cftalg{B}}}{\cftalgf{\cftalg{B}}}
  \ar{d}{\cftctxext[\cftalg{B}]}
  \\
\cftalgc{\cftalg{A}}
  \ar{r}[swap]{\cfthomc{\cfthom{f}}}
& \cftalgc{\cftalg{B}}
\end{tikzcd}
\end{equation}
and
\begin{equation}\label{eq:exthom2}
\begin{tikzcd}[column sep=huge]
\ctxext{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext[\cftalg{A}]}{\cftalgf{\cftalg{A}}}}
  \ar{r}{\jvcomp{}{\cfthomf{\cfthom{f}}}{\jcomp{}{\cftctxext[\cftalg{A}]}{\cfthomf{\cfthom{f}}}}}
  \ar{d}[swap]{\cftfamext[\cftalg{A}]}
& \jcomp{}{\cfthomc{\cfthom{f}}}{\ctxext{\cftalgf{\cftalg{B}}}{\jcomp{}{\cftctxext[\cftalg{B}]}{\cftalgf{\cftalg{B}}}}}
  \ar{d}{\jcomp{}{\cfthomc{\cfthom{f}}}{\cftfamext[\cftalg{B}]}}
  \\
\cftalgf{\cftalg{A}}
  \ar{r}[swap]{\cfthomf{\cfthom{f}}}
& \jcomp{}{\cfthomc{\cfthom{f}}}{\cftalgf{\cftalg{B}}}
\end{tikzcd}
\end{equation}
commute judgmentally.
\end{defn}

@@

As with ordinary morphisms, extension homomorphism can be considered \emph{over} a
morphism.

\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ be a morphism, let $\extalg{P}$ be an extension
algebra in context $\ctxext{\Gamma}{A}$ and let $\extalg{Q}$ be an extension
algebra in context $\ctxext{\Gamma}{B}$. An \emph{extension homomorphism from
$\cftalg{P}$ to $\cftalg{Q}$ over $f$ in context $\Gamma$} is defined to be
an extension homomorphism from $\extalg{P}$ to $\jcomp{A}{f}{\extalg{Q}}$ in
context $\ctxext{\Gamma}{A}$. 
\end{defn}

\begin{eg}
Let $\cfthom{f}$ be an extension homomorphism from $\extalg{A}$ to $\extalg{B}$
in context $\Gamma$. Then $\boldsymbol{\mathcal{F}}_{\cfthom{f}}$ is an
extension homomorphism from $\extalg{F_A}$ to $\extalg{F_B}$ over
$\cfthomc{\cfthom{f}}$ in context $\Gamma$.
\end{eg}

@@

Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$. Weakening will be a
CFT-homomorphism
\begin{equation}\label{cftwk}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
  {\cftalg{F_{F_A}}}
  {\cftwk{\cftalg{A}}}
\end{equation}
satisfying judgmental equalities which express abstracted analogues of
the rules in \autoref{comp-ww,comp-ew,comp-0w}. The
definition of weakening algebras can be found in \autoref{sec:cftwkalg-defn}.

We begin with some trivial remarks to familiarize ourselves with the situation.
The morphism $\cftwk{\cftalg{A}}$ of \autoref{cftwk} is a triple
$(\cftwkc{\cftalg{A}},\cftwkf{\cftalg{A}},\cftwkt{\cftalg{A}})$ consisting of
the morphisms displayed in the following diagram:
\begin{equation*}
\begin{tikzcd}
\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftalgt{\cftalg{A}}}}
  \ar[fib]{d}
  \ar{r}{\cftwkt{\cftalg{A}}}
& \jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftalgt{\cftalg{A}}}}
  \ar[fib]{d}
  \\
\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftalgf{\cftalg{A}}}}
  \ar[fib]{d}
  \ar{r}{\cftwkf{\cftalg{A}}}
& \jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftalgf{\cftalg{A}}}}
  \ar[fib]{d}
  \\
\ctxwk{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{A}}}
  \ar{r}[swap]{\cftwkc{\cftalg{A}}}
& \jcomp{}{\cftctxext}{\cftalgf{\cftalg{A}}}
\end{tikzcd}
\end{equation*}
In particular, we see that $\cftwk{\cftalg{A}}$ is a CFT-homomorphism from
$\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}$ to $\cftalg{F_A}$ over
$\cftctxext$ in context $\Gamma$.

@@

\subsubsection{Weakening preserves itself}
Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$ and let
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
  {\jcomp{}{\cftctxext}{\cftalg{F_A}}}
  {\cfthom{w}}
\end{equation*}
be a CFT-homomorphism. The abstraction of the rules in \autoref{comp-ww} is the
condition that the diagram
\begin{equation}\label{diag:cftwk-cftwk}
\begin{tikzcd}[column sep=10em]
\ctxwk{\cftalgf{\cftalg{A}}}{{\cftalgf{\cftalg{F_A}}}{\cftalg{F_{F_A}}}}
  \ar{d}
    [swap]{ \ctxwk
        {{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}
        {\boldsymbol{\mathcal{F}}_{\cfthom{w}}}
      }
  \ar{r}{\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cfthom{w}}}}
& \ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_{F_{F_A}}}}
  \ar{d}{\jcomp{}{\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamext}}{\boldsymbol{\mathcal{F}}_{\cfthom{w}}}}
  \\
\ctxwk{{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}{\jcomp{}{\cfthomc{\cfthom{w}}}{\cftalg{F_{F_{F_A}}}}}
  \ar{r}[swap]{%
    \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\cfthom{w}}
                }
            }
        }
    }
& \jcomp{}
    {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamext}}
    {\jcomp{}{\cfthomc{\cfthom{w}}}{\cftalg{F_{F_{F_A}}}}}
\end{tikzcd}
\end{equation}
of CFT-algebras in context 
$ \ctxext
    { { { {\Gamma}
          {\cftalgc{\cftalg{A}}}
          }
        { \cftalgf{\cftalg{A}}}
        }
      { \ctxwk
          {\cftalgf{\cftalg{A}}}
          {\cftalgf{\cftalg{A}}}
        }
      }
    { \ctxwk
        {\cftalgf{\cftalg{A}}}
        {\cftalgf{\cftalg{F_A}}}
      }$ %
commutes. In this diagram, it is not immediately obvious that 
\begin{enumerate}
\item the domain of the CFT-homomorphism
$\jcomp{}{\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamext}}{\boldsymbol{\mathcal{F}}_{\cfthom{w}}}$
is the indicated domain. Note that we have the judgmental equalities
\begin{align*}
\jcomp{}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamext}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}
& \jdeq
\ctxwk
  {\cftalgf{\cftalg{A}}}
  {\jcomp{}{\cftfamext}{\cftalg{F_{F_A}}}}
  \\
& \jdeq
\ctxwk
  {\cftalgf{\cftalg{A}}}
  {\jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\cftalg{F_{F_A}}}}
  \\
& \jdeq
\ctxwk
  {\cftalgf{\cftalg{A}}}
  {\cftalg{F_{F_{F_A}}}}.
\end{align*}
\item the domain and codomain of the CFT-homomorphism
\begin{equation*}
\jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\cfthom{w}}
                }
            }
        }
\end{equation*}
from
$ \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
                }
            }
        }
$ to
$ \jcomp{}
      { \cfthomf{\cfthom{w}}}
      { \jcomp{}
          {\cfthomc{\cfthom{w}}}
          { \jcomp{}
              {\cftctxext}
              { \jcomp{}
                  {\cftctxext}
                  {\cftalg{F_{F_A}}}
                }
            }
        }
$ are indeed judgmentally equal to the indicated domain and codomain. To see
this, we have the following lemma:
\end{enumerate}

\begin{lem}
The following three inference rules are valid
\end{lem}

@@

\subsubsection{Substitution is compatible with itself}
